Inferring Fine-Scale Animal Behaviors using Hidden Markov Models

### 2020-06-14 Report/Sections/results.tex
Many Updates
include CIs

\section{Case study- Fluking in Killer Whales}

@@

To demonstrate this methodology as applied to real-world data, the CarHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada. The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 hertz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 3.1 meters in depth and lasts for at least 10 seconds. Finally, for computational purposes, only dives that reached a depth of at least 20 meters were included in the analysis, resulting a total of 7 dives. Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.

The accelerometer recorded both the acceleration due to animal movement and due to gravity. The magnitude of acceleration due only to animal movement is the overall dynamic body acceleration (ODBA) of the killer whale and is known to be a good proxy of energy expenditure \cite{Gleiss:2011}. However, because accelerometer's orientation is unclear, the average acceleration for each component over the entire data collection window was subtracted out (to account for acceleration due to gravity) and the magnitude of the resulting vector was taken as an estimate for the ODBA. This ad-hoc method of removing the effect of gravity is highly prone to errors, especially because the killer whale changes its orientation often. Future work should use alternative methods to find the overall dynamic body acceleration of the whale.

Velocity in the vertical direction was estimated using a central finite difference of the depth data:
%
$$v_t = \frac{d_{t+1} - d_{t-1}}{2\Delta t} = \frac{d_{t+1} - d_{t-1}}{1/25} $$
%
where $v_t$ is measured in meters per second. Due to the high frequency of data collection, small observation errors in depth readings resulted in relatively large errors in the velocity data. To deal with this, the velocity data was smoothed using a Gaussian kernel function with a width of 51 data points ($\approx$ 1 second) ranging from -3 to 3 standard deviations. In particular, if $\phi$ is the pdf of a standard normal distribution, then:
%
\begin{align*}
v^{(smoothed)}_t &= \sum_{i=-25}^{25} \omega_i v_{t+i} \\
\omega_i &= \frac{\phi\left(\frac{3i}{25}\right)}{\sum_{j=-25}^{25}\phi\left(\frac{3j}{25}\right)}
\end{align*}
%
Figure (\ref{fig:smoothing}) shows the velocity data before and after smoothing.

\begin{figure}[h!]
	\centering
	\includegraphics[height=7.5in]{../Plots/smoothed_v.png}
	\caption{From top to bottom: dive depth, unsmoothed vertical velocity, smoothed vertical velocity, ODBA, and the sum of low-frequency Fourier coefficeints of the ODBA for a randomly selected dive of a killer whale.}
	\label{fig:smoothing}
\end{figure}

Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval. Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval. Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken. To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored. One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation. The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find. An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.

@@

In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation. While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.

@@

\subsection{CarHMM adapted to dive data}

Let $v_t$ represent the velocity in the $z-$direction (derived from depth data) at the end of time interval $t$ and $\tilde a_t$ represent the ODBA Fourier sum of interval $t$. The CarHMM was adapted to the dive data using the following model:

\begin{align*}
	v_{t} &\sim \mathcal{N}\left(\phi_{b_t} v_{t-1} + (1-\phi_{b_t}) \mu_{RL,b_t}, \sigma_{b_t}^2\right) \\
	%
	\tilde a_t &\sim \Gamma\left(\tilde \mu_{b_t}, \tilde \sigma^2_{b_t}\right)
\end{align*}

Note that the arguments in the gamma distribution above correspond to the mean and variance of the distribution, \textit{not} the standard scale and shape parameters. In total, 3 behavioral states were chosen for exploratory analysis as an attempt to capture behavioral states with very negative velocities, very positive velocities, and velocities close to zero. This results in a total of $3^2 + 4*3 = 21$ parameters, all of which were estimated using direct likelihood maximization.

>>>>

\section{Results}

@@

To demonstrate this methodology as applied to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada. The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 hertz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.

Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. We set $\tilde{f} = 5$ hertz and $w = 2$ seconds to calculate $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$. This reduces the dimension of each window from $50 1/s * 2 s = 100$ to 2.

@@

In order to find if the CarHHMM was appropriate for this data, a lag plot was made for both $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that $Z^{*(1)}_{t,t^*}$ exhibits a large degree of autocorrelation. While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, so autocorrelation was not incorporated in the emission distribution of $Z^{*(2)}_{t,t^*}$ for computational purposes.

@@
